VISUAL FEATURES OF INSCRIPTIONS. An issue for EDB (and EAGLE)

### 2016-04-14 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
Lorna Hughes' keynote paper partly converted to LaTeX (not finished!)

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.12.14) and
edition in ICVR, VII 19820.}
\label{fig:3}
\end{figure}

@@

\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.8.3) and edition in
ICVR, VIII 22407.}
\label{fig:4}
\end{figure}

@@

The relative high occurrence of images in Christian inscriptions drives EDB team to try to build a search-by-image,
using photos and drawings, but also tagging them with standardized labels. A “high definition” analysis of non-verbal
language of the inscriptions by Christians of Rome in Late Antiquity surely needs photos, drawings, and so on, but
mostly needs a logical, structured, hierarchically ordered taxonomy of all non textual elements defined by controlled,
firm and shared vocabulary: a \emph{thesaurus imaginum}\footnote{ Surely the experiences of other projects can be useful to
this aim: I think for instance to the solutions presented during the VIth EAGLE International Event in Bari by the
lecture offered by Rebecca Benefiel and Holly Sypniewski about the \emph{Ancient Graffiti of Herculaneum project}.}, that can
be a suitable tool also to trail the activity of different stonecutters’ workshops that served – in the case of EDB
inscriptions - the various users of the Roman Christian catacombs.

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}
\caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
``useless'' caption  \emph{s(an)c(t)a Agnes}, age of pope Honorius I, 625/638. (photo: A. E. Felle).}
\label{fig:8}
\end{figure}

@@

\begin{quotation}
John 10.7: \begin{otherlanguage*}{greek}
Εἶπεν οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι ἐγώ εἰμι ἡ θύρα τῶν προβάτων.
\end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});

John 10, 9: \begin{otherlanguage*}{greek}
ἐγώ εἰμι ἡ θύρα· δι' ἐμοῦ ἐάν τις εἰσέλθῃ σωθήσεται καὶ εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει
\end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
whoever enters through me will be saved.} $^{[a]}$ \emph{They will come in and go out, and find pasture}).

@@

\noindent Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
building. Then, we have to face the issue of the description of the writing not only from the necessary point of
view of palaeography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
notion of the inscriptions in the Islamic world --- where often the letters are also and maybe firstly images (they
are used as decorative friezes, architectural ornaments, figures) and their clarity and readability are not considered
as necessary --- can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).

>>>>

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from \citealp{Manzella1997Felle}, sch. 3.12.14) and
edition in ICVR, VII 19820.}
\label{fig:3}
\end{figure}

@@

\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from \citealp{Manzella1997Felle}, sch. 3.8.3) and edition in
ICVR, VIII 22407.}
\label{fig:4}
\end{figure}

@@

The relative high occurrence of images in Christian inscriptions drives EDB team to try to build a search-by-image,
using photos and drawings, but also tagging them with standardized labels. A “high definition” analysis of non-verbal
language of the inscriptions by Christians of Rome in Late Antiquity surely needs photos, drawings, and so on, but
mostly needs a logical, structured, hierarchically ordered taxonomy of all non textual elements defined by controlled,
firm and shared vocabulary: a \emph{thesaurus imaginum}\footnote{ Surely the experiences of other projects can be useful to
this aim: I think for instance to the solutions presented during the VIth EAGLE International Event in Bari by the
lecture offered by Rebecca Benefiel and Holly Sypniewski about the \emph{Ancient Graffiti of Herculaneum project}; see also in this volume the paper by Rebecca Benefiel}, that can
be a suitable tool also to trail the activity of different stonecutters’ workshops that served – in the case of EDB
inscriptions - the various users of the Roman Christian catacombs.

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}
\caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
``useless'' caption  \emph{s(an)c(t)a Agnes}, age of pope Honorius I. (photo: A. E. Felle).}
\label{fig:8}
\end{figure}

@@

\begin{quotation}
John 10.7: \begin{otherlanguage*}{greek}
\underline{Εἶπεν} οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι \underline{ἐγώ εἰμι ἡ θύρα τῶν προβάτων}.
\end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});

John 10, 9: \begin{otherlanguage*}{greek}
ἐγώ εἰμι ἡ θύρα· \underline{δι' ἐμοῦ ἐάν τις εἰσέλθῃ} σωθήσεται καὶ \underline{εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει}
\end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
whoever enters through me will be saved.} $^{[a]}$ \emph{They will come in and go out, and find pasture}).

@@

\noindent Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
building. Then, we have to face the issue of the description of the writing not only from the necessary point of
view of palaeography (we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
notion of the inscriptions in the Islamic world --- where often the letters are also and maybe firstly images (they
are used as decorative friezes, architectural ornaments, figures) and their clarity and readability are not considered
as necessary --- can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).
